---
title: "Exercise 6 - Non-linear Models and Tree-based Methods"
author: "Jack Blumenau"
output: html_document
---

### Exercise 6.1: Non-linear Models

This question relates to the `College` dataset from the `ISLR` package. Start by loading that package, as well as the `gam` package which will allow us to estimate generalised additive models, the `splines` package (which, unsurprisingly, let's us estimate splines), and the `randomForest` package (guess what that is for?).

```{r, warning=FALSE, message = FALSE}

library(ISLR)
library(gam)
library(splines)
library(randomForest)

```

The `College` data contains several variables for 777 US Colleges in the 1990s. Look at the help file for this data set (`?College`) for a description of the variables that are included. 

In this seminar, we will be experimenting with different approaches to estimating non-linear relationships between the `Expend` variable -- which measures the expenditure per student of each college (in dollars) -- and several other variables in the data.

#### a. Create a series of plots which show the association between the `Expend` variable and the following four predictors: `Outstate`, ` PhD`, `Room.Board`, and `S.F.Ratio`. For which of these variables do you think there is evidence of a non-linear relationship?

```{r}

par(mfrow = c(2,2))
plot(College$Outstate, College$Expend)
plot(College$PhD, College$Expend)
plot(College$Room.Board, College$Expend)
plot(College$S.F.Ratio, College$Expend) 

```

It is a little hard to tell, but there is some evidence here that both `PhD` and `S.F.Ratio` are non-linearly related to the `Expend` variable. For the PhD variable, there is essentially no relationship between the percentage of faculty with a PhD and the expenditure variable for most of the range of X, but a positive relationship between the variables for the 80-100% range. The opposite is true for the student-faculty ratio variable, which is strongly negatively associated with the outcome for lower levels of X but mostly flat for higher levels of X.

#### b. Estimate four regression models, all with `Expend` as the outcome variable, and each including one of the predictors you plotted above. Include a second-degree polynomial transformation of X in each of the models (you can do this by using `poly(x_variable,2)` in the `lm()` model formula). Interpret the significance of the squared term in each of your models. Can you reject the null hypothesis of linearity?

```{r}

summary(lm(Expend ~ poly(Outstate, 2), data = College))
summary(lm(Expend ~ poly(PhD, 2), data = College))
summary(lm(Expend ~ poly(Room.Board, 2), data = College))
summary(lm(Expend ~ poly(S.F.Ratio, 2), data = College))

```

In fact, we can reject the null hypothesis of linearity for *all* of these variables! 


#### c. Using the regression you estimated in part b to describe the association between `Expend` and `PhD`, calculate fitted values across the range of the `PhD` variable. Recreate the plot between these variable that you constructed in part a, and add a line representing these fitted values to the plot (using the `lines()` function) to illustrate the estimated relationship. Interpret the graph. (You will need to use the `predict()` function with the `newdata` argument in order to complete this question. You can also find the range of the `PhD` variable using the `range()` function.) I have given you some starter code below.


```{r, echo = TRUE, eval = FALSE}

fitted_vals_quadratic <- predict(phd_mod_quadratic, newdata = data.frame(PhD = SOME_VALUES_GO_HERE))

plot(SOMETHING_GOES_HERE, SOMETHING_ELSE_GOES_HERE,
     xlab = "Percentage of faculty with a PhD",
     ylab = "Predicted Expenditure per Student")
lines(SOME_VALUES_GO_HERE, SOME_OTHER_VALUES_GO_HERE, col = "red")

```



```{r}

phd_mod_quadratic <- lm(Expend ~ poly(PhD, 2), data = College)
range(College$PhD) # Find the range of the PhD variable
fitted_vals_quadratic <- predict(phd_mod_quadratic, newdata = data.frame(PhD = 8:103))

plot(College$PhD, College$Expend,
     xlab = "Percentage of faculty with a PhD",
     ylab = "Predicted Expenditure per Student")
lines(8:103, fitted_vals_quadratic, col = "red")

```

The graph suggests that predicted expenditure per student decreases when moving from low to moderate percentages of faculty with PhDs, and then increases when moving from moderate to high percentages of faculty with PhDs.

#### d. Re-estimate the `Expend ~ PhD` model, this time including a cubic polynomial (i.e. of degree 3). Can you reject the null hypothesis for the cubic term? Add another line (in a different colour) with the fitted values from this model to the plot you created in part c.

```{r}


phd_mod_cubic <- lm(Expend ~ poly(PhD, 3), data = College)
summary(phd_mod_cubic) # Yes, we can reject the null

fitted_vals_cubic <- predict(phd_mod_cubic, newdata = data.frame(PhD = 8:103))

plot(College$PhD, College$Expend,
     xlab = "Percentage of faculty with a PhD",
     ylab = "Predicted Expenditure per Student")
lines(8:103, fitted_vals_quadratic, col = "red")
lines(8:103, fitted_vals_cubic, col = "green")


```

#### e. Estimate a new model for the relationship between `Expend` and `PhD`, this time using a cubic spline instead of a polynomial. You can implement the cubic spline by using the `bs()` function, which is specified thus: `lm(outcome ~ bs(x_variable, df = ?, degree = 3))`. Select a value for the `df` argument that you think is reasonable. Estimate the model and then, again, plot the fitted values across the range of the `PhD` variable. 

```{r}


phd_mod_spline <- lm(Expend ~ bs(PhD, df = 6, degree = 3), data = College)

fitted_vals_spline <- predict(phd_mod_spline, newdata = data.frame(PhD = 8:103))

plot(College$PhD, College$Expend,
     xlab = "Percentage of faculty with a PhD",
     ylab = "Predicted Expenditure per Student")
lines(8:103, fitted_vals_quadratic, col = "red")
lines(8:103, fitted_vals_cubic, col = "green")
lines(8:103, fitted_vals_spline, col = "blue")


```
I have set `df = 6` which is likely too flexible for this data. Nevertheless, the general pattern is consistent with both the cubic and quadratic models.

#### f. Guess what? Now it's time to do the same thing again, but this time using a `loess()` model. The key parameter here is the `span`. High values for `span` will result in a less flexible model, and low values for `span` will result in a more flexible model. Pick a value that you feel is appropriately wiggly. Again, add it to your (now very colourful) plot.


```{r}


phd_mod_loess <- loess(Expend ~ PhD, data = College, span = .1)

fitted_vals_loess <- predict(phd_mod_loess, newdata = data.frame(PhD = 8:103))

plot(College$PhD, College$Expend,
     xlab = "Percentage of faculty with a PhD",
     ylab = "Predicted Expenditure per Student")
lines(8:103, fitted_vals_quadratic, col = "red")
lines(8:103, fitted_vals_cubic, col = "green")
lines(8:103, fitted_vals_spline, col = "blue")
lines(8:103, fitted_vals_loess, col = "purple")


```
I have definitely over-wiggled here!

#### g. Examine the nice plot you have constructed. Which of the lines of fitted values best characterise the relationship between `PhD` and `Expend` in the data? Can you tell?

No, you cannot tell. It is very hard to work out which relationship is best fitting in this example just by looking at plots! Fortunately, tomorrow we will discuss a set of tool that allows us to work out which relationship is most appropriate by seeing which is best at predicting data out of sample.

#### h. Fit a generalised additive model (GAM) to the `College` data using the `gam()` function. This model is just like the `lm()` function, but it allows you to include flexible transformations of the covariates in the model. In this example, estimate a GAM with `Expend` as the outcome, and use all four of predictors that you plotted in part a as well as the `Private` variable. For each of the continuous predictors, use a cubic spline (`bs()`) with 4 degrees of freedom. Once you have estimated your model, plot the results by passing the estimated model object to the `plot()` function. Interpret the results.

```{r}
library(gam)
gam_mod <-  gam(Expend ~ Private + bs(PhD, df=4) + 
                  bs(Outstate, df=4) + bs(Room.Board, df=4) + 
                  bs(S.F.Ratio, df=4) ,
                data=College)

par(mfrow=c(2, 3))
plot(gam_mod, se=TRUE, col="blue")
```

### Exercise 6.2: Tree-based methods

In this question you will use data on UK parliamentary constituencies to predict the vote share won by the Conservative Party in the 2019 general election. Load the data using the following code:

```{r, echo = TRUE}

bes19 <- read.csv("https://raw.githubusercontent.com/lse-me314/assignment06/master/bes19.csv")

```

The `bes19` data contains 631 rows and 16 columns. Each row represents a parliamentary consituency in Great Britain, and the columns include information about each of these constituencies. We are interested in predicting the `Con19` column, which indicates the percentage of the popular vote won by the Conservative candidate in the relevant constituency. 

The other variables in this data are measures taken from the census, which measure a range of factors including the fraction of the constituency's population that was born in England, the fraction with higher-level qualifications, the fraction long term unemployed, the religious composition of the constituency, and so on. We will use these covariates to predict the Conservative vote share in each seat.


#### a. For this task, you should fit the models on a randomly selected subset of your data -- the training set -- and evaluate their performance on the rest of the data -- the test set. Use the code below to create the train- and test-set. I have written the code below to help set up the data for use in this task. Your job in this question is to use some R comments to describe what every line of code does.

```{r}

set.seed(0112358) # Set a value for the random number generator to make the results comparable across runs (extra credit if you know why these digits were selected)
train <-  sample(nrow(bes19), 2/3 * nrow(bes19)) # Randomly select two-thirds of the data
bes19_train <- bes19[train,] # Subset to the training observations
bes19_test <- bes19[-train,] # Subset to the test observations

```

#### b. Train a linear regression model using the `bes19_train` data, predicting `Con19` as a function of all of the other variables in the data. Use this model to generate fitted values (use the `predict()` function) for the **test** data (`bes19_test`).  Create a plot which compares these fitted values to the true values for `Con19` for the test data.

```{r}

# Estimate the linear regression model
lm.con19 <-  lm(Con19 ~  ., 
                data=bes19_train)

# Construct fitted values

bes19_test$Con19_lm_pred <- predict(lm.con19, newdata = bes19_test)

# Create plot

plot(x = bes19_test$Con19_lm_pred, 
     y = bes19_test$Con19, 
     xlab = "Linear Regression Predicted", 
     ylab = "Con19 Vote Share", 
     pch = 19, col = "orange")

```


#### b. Now repeat this prediction task, but this time use a bagging model as implemented using the `randomForest()` function from the `library(randomForest)` package. Again, you should train your model on the `bes19_train` data, and predict for the `bes19_test` data (you can calculate predicted values from the bagging model using the `predict()` function that you used in the previous question). 

*Note:* There are four important arguments that you need to include in the `randomForest()` function:

1. `formula` -- the formula for the model you wish to estimate, indicating the relevant outcome variable and the relevant predictors. Here, your outcome variable is `Con19` and you want to include all predictors, so you can just use `Con19 ~ .` for this argument.
2. `data` -- the dataset you wish to use to estimate the model.
3. `mtry` -- the number of variables randomly sampled at each split of the tree. Remember, for bagging this should be equal to the total number of predictors, while for a random forest model it should be a lower number.
4. `na.action` -- this argument deals with missing data. Set it to `na.action = na.omit` to omit any observation with incomplete data.

```{r}

bag.con19 <-  randomForest(Con19 ~  ., 
                            data = bes19_train, 
                            mtry=15, 
                            na.action = na.omit)

bes19_test$Con19_bag_pred <- predict(bag.con19, newdata = bes19_test)


plot(bes19_test$Con19_bag_pred, 
     bes19_test$Con19, 
     xlab = "Bagging Predicted", 
     ylab = "Con19 Vote Share", 
     pch = 19, col = "black")

```


#### c. Repeat this prediction task for a final time, this time using a randomForest model. As for the bagging approach, you can use the `randomForest()` function, but you will need to make a different choice for the `mtry` argument (remember that for a random forest model a good rule of thumb is $m \approx \sqrt{p}$). You should once more train the model on the training data, and predict for the test-set data.



```{r}

rf.con19 <-  randomForest(Con19 ~  ., 
                            data = bes19_train, 
                            mtry=4, # sqrt(15) = 3.9 
                            na.action = na.omit)


bes19_test$Con19_rf_pred <- predict(rf.con19, newdata = bes19_test)


plot(bes19_test$Con19_rf_pred, 
     bes19_test$Con19, 
     xlab = "Random Forest Predicted", 
     ylab = "Con19 Vote Share", 
     pch = 19, col = "darkgreen")

```


#### d. Which of these models does the best job of predicting Conservative vote share out of sample? Use the function below to calculate the root-mean-squared error (RMSE) for each set of test-set predictions. Report this quantity for the linear model, the bagging model, and the random forest model and report the best performing model according to this metric.

```{r}

rmse <- function(predicted, actual) sqrt(mean((predicted - actual)^2))

```

```{r}


rmse(bes19_test$Con19_lm_pred, bes19_test$Con19)
rmse(bes19_test$Con19_bag_pred, bes19_test$Con19)
rmse(bes19_test$Con19_rf_pred, bes19_test$Con19)

```

**The random forest model narrowly outperforms the other two in this instance.**
